---
title: "1 - pre-processing"
author: "DB"
date: "2/11/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

** IMPORTANT WHEN USING DATA FROM kallisto bustools **
Before running this, make sure that gene names have been changed from ENSEMBLE IDs
to gene names (hgnc or mgi).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BUSpaRse)
library(Matrix)
library(DropletUtils)
library(patchwork)
library(tidyverse)
theme_set(theme_bw())
data_dir <- ("../kb")
out_dir <- ("results/QC/")
object_dir <- "objects"
if (!file.exists(out_dir)) {
  dir.create(out_dir, recursive = T)
}
if (!file.exists(object_dir)) {
  dir.create(object_dir)
}
# extract sample names from kallisto bustools output
sample_key <- "D2"
samples <- list.dirs(data_dir, recursive = F, full.names = F)
samples <- samples[grep(sample_key, samples)]
samples <- setNames(samples, samples)
```


```{r read data}
mat_l <- map(samples, ~read_count_output(
  file.path(data_dir, ., "counts_unfiltered"),
  "cells_x_genes",
  tcc = FALSE)
  )
nsamples <- length(mat_l)
```

for knee_plot see:
https://rdrr.io/bioc/BUSpaRse/man/knee_plot.html

```{r knee plots}
df_kp_l <- imap(mat_l, function(mat, y) {
  df_kp <- tibble(total = colSums(mat)) %>%
    arrange(desc(total)) %>%
    rowid_to_column(var = "rank")
})

p <- imap(df_kp_l, function(df_kp, y) {
  ggplot(df_kp, aes(total, rank)) +
    geom_path() +
    scale_x_log10() + scale_y_log10() + annotation_logticks() +
    labs(x = "Total UMI count", y = "Barcode rank") + 
    ggtitle(samples[y])
  })
ncols <- 2
ggsave(file.path(out_dir, paste0("KneePlot.pdf")), 
       wrap_plots(p, ncol = ncols),
       width =11, height = 5 * ceiling(nsamples/ncols)
       )
```


DropletUtils parameters:
ignore: minimum number of UMI for a real cell, DEFAULT = NULL
lower: DEFAULT = 100

```{r find valid cells with DropletUtils}
# Parameters
version <- 1
niters <- 100000 # DEFAULT: 10000
FDR_thr <- 0.0005
lower <- 100
par_l <- list("D22_ileum" = c("FDR_thr" = FDR_thr, "lower" = lower),
              "D22_rectum" = c("FDR_thr" = FDR_thr, "lower" = lower),
              "D23_ileum" = c("FDR_thr" = FDR_thr, "lower" = lower),
              "D23_rectum" = c("FDR_thr" = FDR_thr, "lower" = lower),
              "D24_ileum" = c("FDR_thr" = FDR_thr, "lower" = lower),
              "D24_rectum" = c("FDR_thr" = FDR_thr, "lower" = 500),
              "D25_ileum" = c("FDR_thr" = FDR_thr, "lower" = lower),
              "D25_rectum" = c("FDR_thr" = FDR_thr, "lower" = lower),
              "D26_ileum" = c("FDR_thr" = FDR_thr, "lower" = lower),
              "D26_rectum" = c("FDR_thr" = FDR_thr, "lower" = lower),
              "D27_ileum" = c("FDR_thr" = FDR_thr, "lower" = lower),
              "D27_rectum" = c("FDR_thr" = FDR_thr, "lower" = lower)
              )
set.seed(100000)
options(repr.plot.width=9, repr.plot.height=6)

pdf(file.path(out_dir, paste0("EmptyDrops_v.", version, ".pdf")), 
    width = 10.5, height = 8)
filt_data <- imap(mat_l, function(mat, y) {
  pars <- par_l[[y]]
  lower = pars["lower"]
  FDR_thr = pars["FDR_thr"]
  # Find empty droplets with DropletUtils
  cat("\n## Sample:", y, "\n")
  e_out <- emptyDrops(mat, BPPARAM=BiocParallel::MulticoreParam(), 
                      niters,
                      lower = lower
                      )
  is_cell <- e_out$FDR <= FDR_thr
  cell_pos <- sum(is_cell, na.rm=TRUE)
  cat("Cells:", cell_pos, "\n\n")
  print(table(Limited=e_out$Limited, Significant=is_cell))
  e_out$FDR[is.na(e_out$FDR)] <- 1

  # Plot of total count against the negative log-probability
  df <- data.frame("Total" = e_out$Total, 
                   "LogProb" = -e_out$LogProb, 
                   "Cell" = is_cell)
  df <- df[!is.na(df$Cell), ]
  du_p <- ggplot(df, aes(Total, LogProb, color = Cell)) + 
    geom_point(show.legend = FALSE) +
    scale_x_log10() + annotation_logticks(sides = "b") +
    labs(x = "Total UMI count", y = "-Log Probability")

  print(du_p + plot_annotation(
    title = samples[y], 
    subtitle = paste0("Cells: ", cell_pos, 
                      ", FDR: ", FDR_thr,
                      ", lower: ", lower
                      )
    )
  )
  
  return(e_out$FDR <= FDR_thr)
})
dev.off()
```

## Sample: D22_ileum 
Cells: 15009 

       Significant
Limited FALSE  TRUE
  FALSE 20286   391
  TRUE      0 14618

## Sample: D22_rectum 
Cells: 10887 

       Significant
Limited FALSE  TRUE
  FALSE 40856     0
  TRUE      0 10887

## Sample: D23_ileum 
Cells: 13509 

       Significant
Limited FALSE  TRUE
  FALSE  8356    88
  TRUE      0 13421

## Sample: D23_rectum 
Cells: 12672 

       Significant
Limited FALSE  TRUE
  FALSE  8075   217
  TRUE      0 12455

## Sample: D24_ileum 
Cells: 15136 

       Significant
Limited FALSE  TRUE
  FALSE  7519   268
  TRUE      0 14868

## Sample: D24_rectum 
Cells: 10748 

       Significant
Limited FALSE  TRUE
  FALSE  5003   314
  TRUE      0 10434

## Sample: D25_ileum 
Cells: 10884 

       Significant
Limited FALSE  TRUE
  FALSE  7554    79
  TRUE      0 10805

## Sample: D25_rectum 
Cells: 12125 

       Significant
Limited FALSE  TRUE
  FALSE 15671    68
  TRUE      0 12057

## Sample: D26_ileum 
Cells: 21982 

       Significant
Limited FALSE  TRUE
  FALSE 10102   397
  TRUE      0 21585

## Sample: D26_rectum 
Cells: 14319 

       Significant
Limited FALSE  TRUE
  FALSE 27614     0
  TRUE      0 14319

## Sample: D27_ileum 
Cells: 12597 

       Significant
Limited FALSE  TRUE
  FALSE  7545   162
  TRUE      0 12435

## Sample: D27_rectum 
Cells: 10142 

       Significant
Limited FALSE  TRUE
  FALSE 16863     0
  TRUE      0 10142


```{r FILTER matrix: use emptyDrops results OR knee plot to set threshold}
mat_l_filt <- map2(mat_l, filt_data, ~ .x[, .y])
```

filter cells that express too few UMI (RNA counts)
```{r distribution of number of detected UMIs per cell}
p <- imap(mat_l_filt, function(x, y) {
  counts <- data.frame("log_UMI" = log10(colSums(x)))
  p <- ggplot(counts, aes(log_UMI)) +
    geom_histogram(bins = 80) + 
    scale_x_continuous(n.breaks = 14) +
    ggtitle(samples[y])
  })
ggsave(file.path(out_dir, "DropletUtils_UMI_histogram.pdf"), 
       wrap_plots(p, ncol = 1),
       width = 8, height = 3*nsamples, units = "in")
```

```{r  set filter after inspecting UMI histogram}
# set the exponent of the UMI filter (read directly from the UMI histogram)
log_umi_thr <- 2.75
umi_filt <- set_names(rep(log_umi_thr, nsamples), nm = names(mat_l))
# customize threshold for individual samples if needed
umi_filt["D23_ileum"] <- 2.6
umi_filt["D23_rectum"] <- 2.8
umi_filt["D24_ileum"] <- 2.5
umi_filt["D25_ileum"] <- 2.85
umi_filt["D25_rectum"] <- 2.6
umi_filt["D26_ileum"] <- 2.85
umi_filt["D27_ileum"] <- 3.0
umi_filt["D27_rectum"] <- 2.9
umi_filt <- 10**umi_filt
umi_filt

mat_l_filt2 <- map2(mat_l_filt, umi_filt, ~ .x[, colSums(.x) > .y])
p <- imap(mat_l_filt2, function(x, y) {
  counts <- data.frame("log_UMI" = log10(colSums(x)))
  p <- ggplot(counts, aes(log_UMI)) +
    geom_histogram(bins = 80) + scale_x_continuous(n.breaks = 14) +
    ggtitle(samples[y])
  })
ggsave(file.path(out_dir, "DropletUtils_UMI_histogram_filt.pdf"), 
       wrap_plots(p, ncol = 1),
       width = 8, height = 3*nsamples, units = "in")
```

filter cells that express too few or too many genes (features counts)

the p vector of each dcgMatrix contains the incremental sum of the number of 
non-empty rows in each column (ie number of genes per cell)
```{r distribution of number of detected genes per cell}
breaks <- 10
tx_size <- 9
ncol <- 1

gene_counts <- map(mat_l_filt2, function(x) {
  l <- length(x@p)
  x@p[2:l] - x@p[1:l-1]
})

p <- imap(gene_counts, function(x, y) {
  x <- data.frame("genes_per_cell" = x)
  ggplot(x, aes(genes_per_cell)) +
    geom_histogram(bins = 80) + 
    scale_x_continuous(n.breaks = breaks) +
    theme(text = element_text(size=tx_size)) +
    ggtitle(y)
  })  
ggsave(file.path(out_dir, "DropletUtils_genes_histogram.pdf"), 
       wrap_plots(p, ncol = ncol),
       width = 8, height = 3*nsamples, units = "in")
```

```{r set filter after inspecting gene count histogram}
filt_low <- 0
gene_filt_low <- set_names(rep(filt_low, nsamples), nm = names(mat_l))
gene_filt_low

filt_hi <- 7000
gene_filt_hi <- set_names(rep(filt_hi, nsamples), nm = names(mat_l))
gene_filt_hi["D24_rectum"] <- 4000
gene_filt_hi["D25_ileum"] <- 8000
gene_filt_hi

mat_l_filt3 <- imap(mat_l_filt2, ~ .x[, gene_counts[[.y]] > gene_filt_low[.y] &
                                        gene_counts[[.y]] < gene_filt_hi[.y]])

gene_counts_filt <- map(mat_l_filt3, function(x) {
  l <- length(x@p)
  x@p[2:l] - x@p[1:l-1]
})

p <- imap(gene_counts_filt, function(x, y) {
  x <- data.frame("genes_per_cell" = x)
  ggplot(x, aes(genes_per_cell)) +
    geom_histogram(bins = 80) + 
    scale_x_continuous(n.breaks = breaks) +
    theme(text = element_text(size=tx_size)) +
    ggtitle(y)
  })  
ggsave(file.path(out_dir, "DropletUtils_genes_histogram_filt.pdf"), 
       wrap_plots(p, ncol = ncol),
       width = 8, height = 3*nsamples, units = "in")
```

```{r remove genes that are not expressed in any cell}
mat_l_filt3 <- map(mat_l_filt3, ~ .x[rowSums(.x) > 0, ])
```

```{r create and save Seurat objects and cell IDs}
# write valid cell IDs for demux
cell_ids <- map(mat_l_filt3, colnames)
iwalk(cell_ids, ~ write.table(.x, file = file.path(object_dir, 
                                                   paste0("cell_ids_", .y, ".txt")),
                              quote = F, row.names = F, col.names = F))

seu_l <- imap(mat_l_filt3, ~Seurat::CreateSeuratObject(.x, project = samples[.y]))
saveRDS(seu_l, file.path(object_dir, "seu_list.after_preprocessing.Rds"))
```

