---
title: "standard processing"
author: "DB"
date: "10/17/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
library(Seurat)
library(SeuratWrappers)
library(openxlsx)
library(sparseMatrixStats)
theme_set(theme_bw())
out_dir <- ("results")
object_dir <- "objects"
if (!file.exists(out_dir)) {
  dir.create(out_dir, recursive = T)
  }
if (!file.exists(object_dir)) {
  dir.create(object_dir)
  }

## functions
# visualize individual metadata levels in DimPlot
PlotIndividualClusters <- function(seu, meta.id, nrows = 3, reduction = "umap"){
  cluster_levels <- levels(as.factor(seu[[meta.id]][, 1]))
  p <- map(cluster_levels, function(x) {
    cells <- colnames(seu)[seu[[meta.id]] == x]
    DimPlot(seu, reduction = reduction, 
            cells.highlight = cells, sizes.highlight = 0.01) + 
      NoLegend() + NoAxes() +
      ggtitle(x) + theme(plot.title = element_text(size = 8, face = "bold"))
  })
  wrap_plots(p, nrow = nrows)
  }
```

```{r inspect Seurat object}
seu_l <- readRDS("objects/seu_list.after_preprocessing.Rds")
# add tissue and HIV status annotation
hiv_dict <- c("D22_rectum" = "HIV-", "D22_ileum" = "HIV-", 
              "D23_rectum" = "HIV-", "D23_ileum" = "HIV-", 
              "D24_rectum" = "HIV+", "D24_ileum" = "HIV+", 
              "D25_rectum" = "HIV-", "D25_ileum" = "HIV-", 
              "D26_rectum" = "HIV+", "D26_ileum" = "HIV+", 
              "D27_rectum" = "HIV+", "D27_ileum" = "HIV+")
seu_l <- imap(seu_l, function(x, y) {
  if (grepl("ileum", y)) {
    x$tissue <- "ileum"
  } else {
    x$tissue <- "rectum" 
  }
  x$hiv <- hiv_dict[as.character(x$orig.ident)]
  return(x)
  })

# pct.mt
seu_l <- map(seu_l, function(x) {
  x[["percent.mt"]] <- PercentageFeatureSet(x, pattern = "^MT-")
  return(x)
})

bins <- 100
breaks <- 12
pdf(file.path(out_dir, "QC", "Histo_QC1_mt.pdf"), width = 10, height = 6)
iwalk(seu_l, function(x, y) {
  metadata <- x[[]]
  p <- ggplot(metadata, aes(x = percent.mt)) + 
    geom_histogram(bins = bins) +
    scale_x_continuous(breaks = scales::breaks_extended(breaks)) +
    ggtitle(paste0(y, " before filtering"))
  print(p)  
  cat(y, "- Genes/Cells:", dim(x), "\n")
  })
dev.off()
```

D22_ileum - Genes/Cells: 27322 12573 
D22_rectum - Genes/Cells: 25734 7984 
D23_ileum - Genes/Cells: 27565 12760 
D23_rectum - Genes/Cells: 26893 10909 
D24_ileum - Genes/Cells: 26872 13704 
D24_rectum - Genes/Cells: 25962 9449 
D25_ileum - Genes/Cells: 27212 9568 
D25_rectum - Genes/Cells: 26811 11176 
D26_ileum - Genes/Cells: 27386 17941 
D26_rectum - Genes/Cells: 26896 12435 
D27_ileum - Genes/Cells: 27356 10405 
D27_rectum - Genes/Cells: 26864 8108 

```{r filter cells with high pct.mt and remove bad samples}
# remove samples without low pct mt cells
bad_samples <- c("D24_rectum")
seu_l <- seu_l[-which(names(seu_l) %in% bad_samples)]
n_samples <- length(seu_l)

pct_mt_cutoff <- 15
pct_mt_filt <- set_names(rep(pct_mt_cutoff, n_samples), nm = names(seu_l))
pct_mt_filt["D22_ileum"] <- 12
pct_mt_filt["D23_ileum"] <- 12
pct_mt_filt["D27_ileum"] <- 12

seu_l <- map2(seu_l, pct_mt_filt, ~ subset(.x, subset = percent.mt < .y))

iwalk(seu_l, ~ cat(.y, "- Genes/Cells:", dim(.x), "\n"))
```

# dataset integration

```{r integration function}
integrate_seu <- function(seuList, nfeatures = 3000) {
  features <- SelectIntegrationFeatures(seuList, nfeatures = 3000)
  seuList <- PrepSCTIntegration(seuList, anchor.features = features)
  seu_int_SCT <- FindIntegrationAnchors(seuList, normalization.method = "SCT", 
                                        anchor.features = features) %>%
    IntegrateData(., normalization.method = "SCT")
  seu_int_SCT <- RunPCA(seu_int_SCT)
}
```


```{r SCT integration of individual tissues: SCTransform}
seu_l <- map(seu_l, SCTransform, variable.features.n = NULL)
seu_tissue_l <- list("ileum" = seu_l[grepl("ileum", names(seu_l))],
                     "rectum" = seu_l[grepl("rectum", names(seu_l))]
                     )
saveRDS(seu_tissue_l, file.path(object_dir, "seu_tissue_list.v2.Rds"))
```


```{r SCT integration of individual tissues: integration}
seu_tissue_int <- map(seu_tissue_l, ~ integrate_seu(.x, nfeatures = 4000))
iwalk(seu_tissue_int, function(x, y) {
  p <- ElbowPlot(x, ndims = 50)
  ggsave(file.path(out_dir, paste0(y, "_ElbowPlot_SCT.pdf")), p)
  })
```

```{r save integrated object}
saveRDS(seu_tissue_int, file.path(object_dir, "seu_tissue_int.v2.Rds"))
```
RELOAD object:
seu_tissue_int <- readRDS("objects/seu_tissue_int.v2.Rds")

```{r SCT integration of individual tissues: cluster & UMAP}
ndims <- 28
seu_tissue_int <- map(seu_tissue_int, function(x) {
  x <- RunUMAP(x, dims = 1:ndims)
  x <- FindNeighbors(x, dims = 1:ndims)
  x <- FindClusters(x) 
  })
```

```{r change cluster resolution}
clus_res <- 0.5
seu_tissue_int <- map(seu_tissue_int, ~ FindClusters(.x, resolution = clus_res))
```


```{r inspect SCT integration}
# orig.ident, seurat_clusters, hiv
plot_id <- setNames(nm = c("orig.ident", "seurat_clusters", "hiv"))
iwalk(seu_tissue_int, function(seu, name) {
  p <- map(plot_id, ~ DimPlot(seu, group.by = .x, shuffle = T, label = T))
  ggsave(file.path(out_dir, 
                   paste0(name, "_tissue.int_DimPlot_QC_clus.res", clus_res, ".pdf")
                   ), 
         width = 8, height = 7*length(plot_id),
         plot = wrap_plots(p, ncol = 1))  
})
# percent mt and UMI counts
iwalk(seu_tissue_int, function(seu, name) {
  p1 <- DimPlot(seu, group.by = "seurat_clusters", shuffle = T, label = T)
  p2 <- FeaturePlot(seu, features = "percent.mt")
  p3 <- FeaturePlot(seu, features = "nCount_RNA", max.cutoff = "q90")
  ggsave(file.path(out_dir, 
                   paste0(name, "_tissue.int_pct.mt_nCount.RNA_clus.res", 
                          clus_res, ".pdf"
                          )
                   ),
         plot = p1 + p2 + p3,
         width = 18, height = 6
         )
  })
# cluster use by sample 
summary <- map(seu_tissue_int, function(seu) {
  seu[[]] %>%
    group_by(orig.ident, seurat_clusters) %>%
    summarize(count = n()) %>%
    pivot_wider(values_from = count, names_from = orig.ident, values_fill = 0)
  }) 
write.xlsx(summary, 
           file.path(out_dir, 
                     paste0("summary_clusters_tissue.int_clus.res", clus_res, ".xlsx")
                     ),
           overwrite = T
           )
```


```{r annotation with marker genes}
# DotPlot
features <- list("stem cells" = c("LGR5","LGR4","OLFM4", "ASCL2", "SOX9",
                                  "GNB2L1", "EEF1B2", "EEF1A1", "C1QBP"),
                 "TA cells" = c("HELLS", "PCNA", "KRT18", "CCND2",
                                "G0S2", "SAA2", "PDZK1IP1"),
                 "secretory TA" = c("TOP2A", "CCNB1", "UBE2S"),
                 "cycling TA" = c("MKI67", "CENPF", "NPM1", "HIST1H4C", "AKAP12"),
                 "enterocytes" = c("ALPI", "FABP1", "KRT19", "KRT20", 
                                   "CA2", "ANPEP", "VIL1"),
                 "enterocytes 1" = c("GUCA2A", "FABP6", "CA4", "CEACAM7", "SLC40A1",
                                     "SLC26A3", "APOA1", "APOB", "RBP2", "ALDOB"),
                 "enterocytes 2" = c("MUC1", "CLCA4", "MUC13", "LAMA3", "ABCB1"),
                 "immature enterocytes 1" = c("INSIG1", "CKB", "NEAT1", "SCD", "PHGR1"),
                 "immature enterocytes 2" = c("PLCG2", "MMP7", "CXCL1", "IL32", 
                                              "CCL2", "SAA1"),
                 "goblet" = c("MUC2", "MUC6", "ZG16", "SPINK4", "ATOH1", "TFF3"),
                 "paneth" = c("LYZ", "PLA2G2A", "DEF6"),
                 "endocrine" = c("CHGA", "CHGB", "REG4", "NEUROD1", "PYY", "NTS", "GCG"),
                 "tuft" = c("DCLK1", "TRPM5", "RGS13", "PTGS1"),
                 "M cells" = c("GP2", "SPIB"),
                 "SARS-CoV-2" = c("ACE2", "TMPRSS2", "TMPRSS4", "DPP4", "CTSL"))

iwalk(seu_tissue_int, function(seu, name) {
  p <- DotPlot(seu, features = features) + 
    RotatedAxis() + 
    theme(text = element_text(size=8), axis.text.x = element_text(size=8))
  ggsave(file.path(out_dir, 
                   paste0(name, "_DotPlot_int_clus.res", clus_res, ".pdf")
                   ),
         plot = p,
         width = 20, height = 6)  
})

# my feature list
features2 <- list("stem" = c("CCL25","LGR5","CDCA7","MYC","OLFM4","RGMB", "ASCL2"),
                  "cc_s" = c( "MCM5","PCNA","RRM1","FEN1","MCM4"),
                  "cc_g2_m" = c("HMGB2","CDK1","NUSAP1","UBE2C","BIRC5"),
                  "immature enterocytes" = c("CKB","NEAT1","PHGR1","PLCG2","CXCL1","CCL2"),
                  "enterocyte" = c("RBP2","ANPEP","DGAT1","FABP2","APOB","KHK",
                                   "APOA4","FABP1", "KRT19", "KRT20"),
                  "goblet" = c("MUC2","CLCA1","SPINK4","BCAS1","ATOH1","ZG16","TFF3"),
                  "paneth" = c("LYZ", "PLA2G2A","DEFA5","DEFA6","REG3A"),
                  "endocrine" = c("OR51E1","CDHR3","KCTD12","RAB3C","CHGB","REG4","NEUROD1","PYY"),
                  "tuft" = c("SH2D6","LRMP","FYB1","PSTPIP2","TRPM5","RGS13","PTGS1"),
                  "SARS-CoV-2" = c("ACE2", "TMPRSS2", "TMPRSS4", "DPP4", "CTSL")
                  )
iwalk(seu_tissue_int, function(seu, name) {
  p <- DotPlot(seu, features = features2) + 
    RotatedAxis() + 
    theme(text = element_text(size=8), axis.text.x = element_text(size=8))
  ggsave(file.path(out_dir, 
                   paste0(name, "_DotPlot_list2_int_clus.res", clus_res, ".pdf")
                   ),
         plot = p,
         width = 20, height = 6)  
})
```
Ileum:
Warning: Found the following features in more than one assay, excluding the default. We will not include these in the final data frame: LGR5, LGR4, SOX9, EEF1B2, EEF1A1, PDZK1IP1, VIL1, CEACAM7, LAMA3, ABCB1, SCD, MMP7, SAA1, MUC6, DCLK1, GP2, ACE2, TMPRSS2, TMPRSS4
Warning: Could not find CCL2 in the default search locations, found in RNA assay instead

Rectum:
Warning: Found the following features in more than one assay, excluding the default. We will not include these in the final data frame: LGR5, LGR4, SOX9, EEF1B2, EEF1A1, C1QBP, CCND2, PDZK1IP1, NPM1, ALPI, VIL1, APOB, RBP2, SCD, MMP7, SAA1, MUC6, CHGA, NTS, DCLK1, GP2, ACE2, TMPRSS4, DPP4, CTSL

# process ileum only
(rectum cell types are positioned in unstandard relationships in the UMAP)

```{r annotate clusters}
clusters_recode <- c("3" = "stem cells",
                     "11" = "TA cells",
                     "10" = "immature enterocytes-1",
                     "12" = "immature enterocytes-2",
                     "7" = "enterocytes",
                     "2" = "enterocytes_1",
                     "20" = "enterocytes-M cells",
                     "0" = "goblet-1",
                     "6" = "goblet-2",
                     "24" = "goblet-3",
                     "8" = "goblet-4",
                     "13" = "goblet-5",
                     "19" = "paneth",
                     "18" = "tuft",
                     "14" = "endocrine-1",
                     "17" = "endocrine-2",
                     "1" = "undefined-1",
                     "5" = "undefined-2",
                     "9" = "undefined-3",
                     "15" = "undefined-4",
                     "4" = "undefined-5",
                     "16" = "undefined-6",
                     "21" = "undefined-7",
                     "22" = "undefined-8",
                     "23" = "undefined-9"
                     )
seu_int_ileum <- seu_tissue_int$ileum
seu_int_ileum$cell.type <- factor(clusters_recode[levels(
  seu_int_ileum$integrated_snn_res.0.5
  )[seu_int_ileum$integrated_snn_res.0.5]],
  levels = clusters_recode)
Idents(seu_int_ileum) <- "cell.type"

DimPlot(seu_int_ileum, group.by = "cell.type", shuffle = T, label = T)
ggsave(file.path(out_dir, "ileum_cell.type.pdf"),
       width = 10, height = 8)

# cell type by HIV status 
summary <- seu_int_ileum[[]] %>%
  group_by(cell.type, hiv) %>%
  summarize(count = n()) %>%
  pivot_wider(values_from = count, names_from = hiv, values_fill = 0)

write.xlsx(summary, 
           file.path(out_dir, 
                     paste0("summary_ileum_cell.type_hiv_clus.res", clus_res, ".xlsx")
                     ),
           overwrite = T
           )
```


```{r differential test: annotation by HIV status}
ids <- levels(seu_int_ileum$cell.type)
names(ids) <- ids
de_results <- map(ids, ~ FindMarkers(seu_int_ileum, 
                                     min.pct = 0.2,
                                     ident.1 = "HIV+", ident.2 = "HIV-", 
                                     assay = "SCT",
                                     group.by = "hiv", 
                                     subset.ident = .x
                                     )
                  )
# save DE results and object
write.xlsx(de_results, 
           file.path(out_dir, paste0("de_results_ileum.xlsx")), 
           rowNames = T
           )
saveRDS(de_results, file.path(object_dir, "de_results_seu_int_ileum.Rds"))

# ROC test
de_results_roc <- map(ids, ~ FindMarkers(seu_int_ileum, 
                                     min.pct = 0.2,
                                     test.use = "roc",
                                     ident.1 = "HIV+", ident.2 = "HIV-", 
                                     assay = "SCT",
                                     group.by = "hiv", 
                                     subset.ident = .x
                                     )
                  )
# save DE results and object
write.xlsx(de_results_roc, 
           file.path(out_dir, paste0("de_results_roc_ileum.xlsx")), 
           rowNames = T
           )
saveRDS(de_results, file.path(object_dir, "de_results_roc_seu_int_ileum.Rds"))
```

Results from Wilcoxon test (default) and ROC are very consistent
-> use Wilcoxon

```{r examine a few differentially expressed genes}
cell_types_use <- set_names(c("stem cells", "enterocytes_1", "goblet-1", "goblet-2"))
DefaultAssay(seu_int_ileum) <- "SCT"
# violin plots
top_n_genes <- 8
feature_list <- map(cell_types_use, ~ rownames(de_results[[.x]])[1:top_n_genes])
iwalk(feature_list, function(feats, name) {
  pdf(file.path(out_dir, paste0("VlnPlt_de_results_ileum_", name, ".pdf")),
      width = 11, height = 8.5
      )
  walk(feats, ~ print(VlnPlot(seu_int_ileum, 
                              features = .x, 
                              group.by = "cell.type", 
                              assay = "SCT",
                              split.by = "hiv", 
                              split.plot = TRUE, 
                              pt.size = 0.1
                              )
                      )
       )
  dev.off()
  })
# feature plots
top_n_genes <- 4
feature_list <- map(cell_types_use, ~ rownames(de_results[[.x]])[1:top_n_genes])
iwalk(feature_list, function(feats, name) {
  range <- rowQuantiles(GetAssayData(seu_int_ileum, assay = "SCT", slot = "data"), 
                        rows = feats,
                        probs = c(0.0, 1),
                        useNames = T)
  # range is duplicated because we are plotting for 2 conditions
  range <- rbind(range, range)  
  range <- as.data.frame(t(range))
  p <- FeaturePlot(seu_int_ileum, 
                   features = feats, 
                   split.by = "hiv",
                   combine = F
                   )
  p <- map2(p, range, function(x, y) {
    cat(y)
    x + scale_color_gradient(low = "lightgrey", high = "blue",
                             limits = c(y[1], y[2])
                             )
    })
  ggsave(file.path(out_dir, paste0("FeaturePlt_de_results_ileum_", name, ".pdf")),
         plot = wrap_plots(p, ncol = 4),
         width = 12, height = 5.5
         )
  })
```

-------------- MNN integration

```{r integrate with fastMNN}
seu_2_mnn <- map(seu_tissue_l, function(x) {
  merge(x[[1]], x[-1]) %>%
  SplitObject()  
})

ndims <- 28
seu_mnn_l <- map(seu_tissue_l, function(x) {
  RunFastMNN(object.list = x,
             assay = "SCT", 
             BPPARAM = BiocParallel::MulticoreParam())
  })


out <- FindNeighbors(out, reduction = "mnn", dims = 1:ndims) %>%
  FindClusters() %>%
  RunUMAP(reduction = "mnn", dims = 1:ndims)
```

